---
title: "Data Exploration for Capstone Project"
author:
  - name: "Jessica Maina"
    orcid: "0009-0007-5229-3182"
    email: "Jessica.Maina@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
embed-resources: true
editor: visual
execute: 
  echo: true
  warning: false
editor_options: 
  chunk_output_type: console
---

# Introduction

The **African Development Corridor Database 2022**, compiled by Jessica P.R. Thorn, Ben Mwangi, and Diego Juffe Bignoli, provides a comprehensive geospatial dataset of planned and existing development corridors across Africa, including roads, railways, pipelines, and ports. Published on Dryad in September 2022 and updated in October 2022, the dataset aims to support research on the environmental, social, and economic impacts of infrastructure expansion on biodiversity, ecosystems, and local communities.

# Methods

## Reading the Data

```{r}
library(tidyverse) # Load R packages
install.packages("janitor")
library(janitor)
library(stringr)
library(ggplot2)

# Read data
my_data <- read_csv(here::here("data/raw/AfricanDevelopmentCorridorDatabase2022.csv"))
```

## Data Exploration Approach

```{r}
glimpse(my_data) # Quick data view

dim(my_data)

corridors_per_country <- my_data |> 
  separate_rows(Country, sep = ";") |>
  mutate(Country = trimws(Country)) |> # Remove extra spaces between the country names
  count(Country, sort = TRUE)

knitr::kable(corridors_per_country, caption = "Table 0: Number of Corridors per Country") # Table showing number of corridors in eacb country. This includes corridors shared among more than one country and, countries with missing data (NA) and missing information (NI).bKenya has the highest number of corridors followed by Tanzania. 

development_type <- my_data |> 
  separate_rows(`Infrastructure/development type`, sep = ";") |>
  mutate(`Infrastructure/development type` = trimws(`Infrastructure/development type`)) |> 
  count(`Infrastructure/development type`, sort = TRUE) # Roads are the most invested in corridors in Africa to drive development in the continent.

```

## Initial Data Tidying

```{r}
corridor_data <- my_data |>
  janitor::clean_names() # clean variable names
# janitor function cleaned the column names by converting them to lowercase and replacing spaces between the words witb underscores

corridor_data |>
  mutate(
    launch_year = as.numeric(launch_year)  
  ) # convert character columns to numeric type for visualisation 


corridor_data <- corridor_data |>
  mutate(
    amount_funded_total = str_extract_all(amount_funded_usd_millions_per_donor_type, "\\d+\\.*\\d*") |> 
      lapply(as.numeric) |> 
      sapply(sum, na.rm = TRUE)
  ) # str_extract_all extracts all numbers from the observations in  amount_funded_usd_mllions_per_data columns. The \\d+\\.*\\d* extracts all numbers numbers while skiping the NI. 
# lapply(as.numeric) selects all observations in the list as individuals and convert them to numeric values.
# sapply(sum, na.rm = TRUE) selects all observations in the list as indviduals and sums them up for each project in a country.  


corridor_data <- corridor_data |>
  mutate(
    distance_km_maximum = as.numeric(distance_km_maximum),
    usd_amount_million_maximum = as.numeric(usd_amount_million_maximum),
    launch_year = as.numeric(launch_year)
  ) # Final conversion of variables from character data to numeric data needed for data summary and visualisation 
```

# Results

```{r}
country_corridor_data <- corridor_data |>
  group_by(country) |>
  summarize(
    n_corridors = n(),
    avg_length_km = mean(distance_km_maximum, na.rm = TRUE),
    avg_usd_amount = mean(usd_amount_million_maximum, na.rm = TRUE),
    latest_launch = max(launch_year, na.rm = TRUE),
    total_usd_funded = sum(amount_funded_total, na.rm = TRUE)
  ) # Summary of the corridor data from 185 rows and 25 columns to 39 rows and 6 columns. 

clean_country_corridor_data <- country_corridor_data |>
  filter(
    if_all(everything(), ~ !is.infinite(.) & !is.nan(.))
  ) # Removed all rows with missing and infinity data (Nan & Inf) for better visualisation of the summary data. 
# if_all() function refers to all columns in a row while everything() refers to all columns in the country_county_data dataset. 

clean_country_corridor_data |>
  arrange(desc(total_usd_funded)) |>
  knitr::kable(digits = 1, caption = "Table 1: Countries by Total USD Funded") # Display a clean table of the data summary. This table presents the summary data, that is, countries, number of corridors, average length, total money used, latest launch year and total money funded. The table is arranged in the order of the most money funded to the least. # From the table, the Benguela Railway corridor shared by Angola and Democratic Republic of Congo received the highest amount of money from donors. 


individual_clean_country_data <- clean_country_corridor_data |>
  separate_rows(country, sep = ";") |>
  mutate(country = str_trim(country)) # Split columns with more than one country to individual countries. This is for easier visualisation for each country and its corridors.


summary_individual_country_data <- individual_clean_country_data |>
  group_by(country) |>
  summarise(
    n_corridors = n(),
    avg_length_km = mean(avg_length_km, na.rm = TRUE),
    avg_usd_amount = mean(avg_usd_amount, na.rm = TRUE),
    latest_launch = max(latest_launch, na.rm = TRUE),
    total_usd_funded = sum(total_usd_funded, na.rm = TRUE),
    .groups = "drop"
  ) # Summary of the columns with multiple corridors of the same country to make one column for each country.

summary_individual_country_data |>
  arrange(desc(total_usd_funded)) |>
  knitr::kable(digits = 1, caption = "Table 2: Individual Countries by Total USD Funded") # A clean table of the final country summary data. This table represents each country in its own row compared to the above with more than one country in a row. This is to ensure that during data visualisation each country is outlined in it own. # From the table, Angola had 2 corridors between 2012 and 2016, which had the highest amount of money funded by donors. The third corridor was shared between Angola and DRC.

ggplot(summary_individual_country_data,
       mapping = aes(x = country, y = total_usd_funded)) +
  geom_boxplot(fill = "gray90", outlier.shape = NA) +
  geom_jitter(width = 0.2, aes(color = country), size = 3) +
  labs(title = "Figure 1: Total USD Funded per Country",
       y = "Total USD Funded",
       x = "Country") +
  theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) # Country names to align vertically to avoid overlapping    # This plot visualizes the distribution of total USD funded per country using a boxplot combined with jittered points for individual observations. The boxplot summarizes the spread, median, and potential outliers (hidden), while the jitter points reveal the variation within each country. Angola has the highest USD funded of 12881.9Usd by donors like Government of Angola. DRC is the second followed by Rwanda and Togo. 

summary_individual_country_data |>
  arrange(desc(total_usd_funded)) |>
  head(10) |>
  ggplot(aes(x = country, y = total_usd_funded)) +
  geom_boxplot(fill = "gray90", outlier.shape = NA) +
  geom_jitter(width = 0.2, aes(color = country), size = 3) +
  labs(title = "Figure 2: Total USD Funded per Country",
       y = "Total USD Funded",
       x = "Country") +
  theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) # smaller plot with the first 10 countries, in descending order, for a clearer reading. Findings are the same as the above plot. 
```

## Methods

```{r}
corridor_data <- corridor_data |>
  mutate(
    status = case_when(
      str_to_lower(status) %in% c("completed", "complete") ~ "completed",
      str_to_lower(status) %in% c("under construction", "in progress") ~ "under construction",
      TRUE ~ status
    )
  ) # lower case names for all variables 

ggplot(summary_individual_country_data,
       mapping = aes(x = country, y = n_corridors, 
               size = avg_length_km, color = avg_length_km)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(3, 10)) +
  scale_color_viridis_c() +
  labs(title = "Figure 3: Number of Corridors and Length by Country",
       x = "Country",
       y = "Number of Corridors",
       size = "Total Length (km)",
       color = "Total Length (km)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Viridis color palette is colour blind friendly. # This scatter plot shows the number of corridors per country on the y-axis, with countries on the x-axis. Each point's size and color intensity represent the average corridor length in kilometers, larger and darker points indicate longer corridors. # Eygpt has the longest average corridor length followed by Sudan and Mauritania.


ggplot(summary_individual_country_data, 
       mapping = aes(x = avg_usd_amount, y = total_usd_funded, color = country)) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Figure 4: Average USD Amount Used vs. Total USD Funded",
       x = "Average USD Amount per Country",
       y = "Total USD Funded",
       color = "Country") +
  theme_minimal() # This scatter plot explores the relationship between the average USD amount per country and the total USD funded.

summary_individual_country_data |>
  arrange(desc(total_usd_funded)) |>
  head(10) |>
  ggplot(aes(x = avg_usd_amount, y = total_usd_funded, color = country)) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Figure 5: Top 10 Countries by Total USD Funded",
       x = "Average USD Amount per Corridor",
       y = "Total USD Funded",
       color = "Country") +
  theme_minimal() # filter to the first 10 countries with the highest amount funded for a clearer reading. From the above two plots, Togo had the most expensive corridor of 4,150 USD followed by  Rwanda with 4,016.5 USD. However, Angola has the highest USD funded of 12881.9 USD followed by DRC.  


summary_individual_country_data |>
  select(country, n_corridors, latest_launch) |>
  arrange(desc(latest_launch)) |>
  knitr::kable(digits = 0, caption = "Table 3: Country Corridors by Latest Launch Year") # Table showing the country corridors in a descending order of the launch years. # Algeria has the latest corridor known as Trans-Maghreb Corridor which was launched in 2024. 

```

# Conclusions

```{r}
write_csv(corridor_data, here::here("data/processed/AfricanDevelopmentCorridorDatabase2022_tidied.csv"))
```

## Summary of Findings

This data cycle has:

-   Loaded and cleaned the data

-   Explored it with key descriptive statistics

-   Fixed several common data issues (character dates, inconsistent categories, entry errors)

## Questions and Next Steps

Next steps will include:

Analyzing the data in terms of regions or development type

Data visualization
