---
title: "Data Exploration for Capstone Project"
author:
  - name: "Jessica Maina"
    orcid: "0009-0007-5229-3182"
    email: "Jessica.Maina@colorado.edu"
    affiliation:
      - name: "University of Colorado Boulder"
        department: "Department of Civil, Environmental and Architectural Engineering"
        city: "Boulder"
        state: "CO"
        country: "USA"
date: today
format: html
editor: visual
---

# Introduction

The **African Development Corridor Database 2022**, compiled by Jessica P.R. Thorn, Ben Mwangi, and Diego Juffe Bignoli, provides a comprehensive geospatial dataset of planned and existing development corridors across Africa, including roads, railways, pipelines, and ports. Published on Dryad in September 2022 and updated in October 2022, the dataset aims to support research on the environmental, social, and economic impacts of infrastructure expansion on biodiversity, ecosystems, and local communities.

# Methods

## Reading the Data

```{r}
library(tidyverse) # Load package

install.packages("janitor")
library(janitor)

# Read data

my_data <- read_csv(here::here("data/raw/AfricanDevelopmentCorridorDatabase2022.csv"))
```

## Data Exploration Approach

## Initial Data Tidying

# Results

```{r}
glimpse(my_data)

dim(my_data)

corridor_data <- my_data |>
  janitor::clean_names() # clean variable names


corridor_data_outline <- corridor_data |>
  group_by(country) |>
  summarize(
    n_corridors = n(),
    avg_length_km = mean(distance_km_maximum, na.rm = TRUE),
    avg_usd_amount = mean(usd_amount_million_maximum, na.rm = TRUE),
    latest_launch = max(launch_year, na.rm = TRUE)
  )

names(corridor_data)

```

## Methods

# Conclusions

## Summary of Findings

## Questions and Next Steps
